---
title: "231B Section 2"
author: "Guadalupe Tuñón \\ guadalupe.tunon@berkeley.edu"
date: "January 28, 2016"
output:
  slidy_presentation:
      fig_height: 4
---

Today
=====================================================================
**1. Sampling with vs without replacement: Variance of the sample mean**

**2. The bootstrap** 

2a. CLT: Distribution of the bootstap mean as a function of sample size

**3. Standard error for the difference in means**

**4. Hypothesis testing**

4a. T-test

4b. Randomization inference


```{r, echo=FALSE}

rm(list=ls())

# Loading needed libraries
library(foreign) # to import csv
library(ggplot2) # for plots using ggplot



```

---

1. Sampling with vs without replacement: the variance of the sample mean
=======================================================================

Last week, we talked about the expected value of the mean and wrote 
a simulation to show that it is an unbiased estimate for the population 
mean. Is this true if we are sampling with replacement? Without?

In lecture on Tuesday, we talked about the variance of the sample mean 
sampling with and without replacement.

> - Are they the same? If not, how does one compare to the other?

> - How could we write a simulation to answer this question?


---

First we need a box to sample from. We will create two different boxes, one with some "Bill Gates" in it.


```{r}

pop1 <- rnorm(50, 0, 5) # no outliers
pop2 <- c(rnorm(48, 0, 5), 8, 11) # two outliers

```

Let's find the average of each box
```{r}
apply(cbind(pop1, pop2), 2, FUN=mean)
```

> - We will sample 25 units from each population with and without replacement 
> - We will repeat this process 10,000 times

---

```{r}

sample_means <- matrix(NA, 10000, 4)

for (i in 1:10000){
  sample_means[i,] <- c(mean(sample(pop1, 25, replace=T)),
                        mean(sample(pop1, 25, replace=F)),
                        mean(sample(pop2, 25, replace=T)), 
                        mean(sample(pop2, 25, replace=F)))
}

head(sample_means)

```

---

```{r, echo = FALSE, fig.height = 4}
sample_means <- as.data.frame(sample_means)
names(sample_means) <- c("pop1_with", "pop1_without",
                         "pop2_with", "pop2_without")

par(mfrow=c(1,2))
# plots for population 1
plot(density(sample_means$pop1_without), lwd=3, col="blue",
     xlim=c(-5, 5),
     main="Distribution of sample means \n population 1 (no outliers)")
lines(density(sample_means$pop1_with), lwd=3, col="red")
# plots for population 2
plot(density(sample_means$pop2_without), lwd=3, col="blue",
     xlim=c(-5, 5),
     main="Distribution of sample means \n population 2 (with outliers)")
lines(density(sample_means$pop2_with), lwd=3, col="red")

```

Blue = sampling without replacement, Red = sampling with replacement

```{r}
apply(sample_means, 2, FUN=mean)
apply(sample_means, 2, FUN=var)

```


2. The bootstrap
==============================================================

```{r, echo = FALSE}

# loading the data - Dunning & Harrison (2010)
setwd("~/Dropbox/231B_Spring2015/Replication files/Dunning_and_Harrison")
cc_data <- read.csv("Cross_Cutting_APSR_ReplicationData.csv")


# treat_assign takes on a value 1 through 6 and denotes the treatment condition to 
# which the respondent was assigned, as follows:
# 1 -- Same ethnicity, joking cousin
# 2 -- Same ethnicity, not joking cousin
# 3 -- Different ethnicity, joking cousin
# 4 -- Different ethnicity, not joking cousin
# 5 -- No last name given for candidate
# 6 -- Candidate and subject have same last name (and thus ethnicity)


```


For this problem, we are going to use the sample values for the co-ethnic cousin 
(treatment) condition (treat_assign==1). This will be our box, the "bootstrap population". 

Let's plot the empirical distribution of responses in the treatment group.

```{r, echo = FALSE}

box <- cc_data$vote_prefer[cc_data$treat_assign==1]
box <- as.data.frame(box)

m <- ggplot(box, aes(x=box))
m + geom_histogram(aes(y = ..density..), alpha=.25, binwidth=1) + 
  # We will overlay a blue line showing the normal distribution with mean equal to the mean
  # of the box and sd equal to the sd of the box
  stat_function(fun=dnorm, args=list(mean=mean(box$box), sd=sd(box$box)), 
                col="blue", size=.8) +
  # and a red line showing the mean of the box
  geom_vline(x=mean(box$box), col="red", size=1.25) +
  theme_bw() + 
  labs(title="",  x="Respondent wants to vote for candidate (1-7 scale)",
       y="Density")

```

---

We will use the bootstrap to investigate the properties of different sampling procedures.

> 1. Take the sample values as the population; 
> 2. Draw a sample from this population (box) with replacement, using the sampling 
procedure we want to analyze; save the sample statistics (e.g. the mean). This is a "bootstrap replicate."
> 3. Repeat step (2) many times (say, 10,000 times).  
> 4. Plot the distribution of the saved statistics across all the bootstrap replicates.
This gives us a good glimpse of the sampling distribution of the statistic of interest.

--- 

```{r}

# We first need to define the number of units we will draw from the box 
# (with replacement). 
# For the first example, let's take N=5
N <- 5

# If we wanted to do step (2) once, we would sample from the box N times with replacement
boot_sample <- sample(box$box, N, replace=T) 
# And then take the statistic of interest for this bootstrap sample, here the mean. 
mean(boot_sample)

```

But we want to do this many times--- 10,000 times! What can we do?

---

We write a for loop that repeats this sampling procedure 10000 times:

```{r}
boot_reps <- 10000 # number of bootstrap replicates we will repeat step (2) for
boot_mean <- NA # placeholder vector for the results

for (i in 1:boot_reps){
  boot_sample <- sample(box$box, N, replace=T) 
  boot_mean[i] <- mean(boot_sample)
}

```

```{r, echo=FALSE, fig.height=3}
# Now we plot the results
m <- ggplot(as.data.frame(boot_mean), aes(x=boot_mean))
# First we plot a histogram with the results
m + geom_histogram(aes(y = ..density..), alpha=.5, binwidth=.1) + 
  # and overlay a line with the density of a normal distribution with mean equal to the 
  # mean of the bootstrap means and sd equal to the sd of the bootstrap means.
  stat_function(fun=dnorm, 
                args=list(mean=mean(boot_mean), sd=sd(boot_mean)), 
                col="blue", size=1) +
  # and we add a vertical line for the mean of the box
  geom_vline(x=mean(box$box), col="red", size=1) + theme_bw()

```

---

We will now repeat this procedure, varying the size of the bootstrap sample, several times. We can write a function which does the bootstraps, and plots a histogram for their mean.

Our function will require 
- the data, 
- the number of bootstrap replicates, 
- the number of observations to be sampled from the box in each replicate
- the binwidth to be used for the histogram. 

It will sample N units with replacement and get the mean for that sample

---

```{r}
bootstrap_mean <- function(data, replicates=10000, N, bin, title="", 
                           xlab="", ylab="", xmax, xmin){
  
  boot_mean <- NA #plaaceholder vector
  for (i in 1:replicates){
    # we sample N units from the box with replacement
    boot_sample <- sample(data, N, replace=T)
    # and save their mean
    boot_mean[i] <- mean(boot_sample)
  }
  
  m <- ggplot(as.data.frame(boot_mean), aes(x=boot_mean))
  # First we plot a histogram with the results
  m + geom_histogram(aes(y = ..density..), alpha=.5, binwidth=bin) + 
    # and overlay a line with the density of a normal distribution with mean equal to the mean
    # of the bootstrap means and sd equal to the sd of the bootstrap means.
    stat_function(fun=dnorm, 
                  args=list(mean=mean(boot_mean), sd=sd(boot_mean)), col="blue", size=1) +
    # and we add a vertical line for the mean of the box
    geom_vline(x=mean(box$box), col="red", size=1) +
    # limits for x axis
    scale_x_continuous(limits = c(xmin, xmax)) +
    # and labels
    labs(title=title,  x=xlab, y=ylab) + theme_bw()
}

```

---

Now we will run the bootstrap varying N.

---

**N = 5**
```{r, fig.height=4}
bootstrap_mean(data=box$box, replicates=10000, N=5, bin=.035, 
               # title="N=5", 
               xmin=1, xmax=7,
               xlab="Mean - Respondent wants to vote for candidate (1-7 scale)", 
               ylab="Density")
```

---

**N = 25**
```{r, fig.height=4}
bootstrap_mean(data=box$box, replicates=10000, N=25, bin=.035, 
               # title="N=25", 
               xmin=1, xmax=7,
               xlab="Mean - Respondent wants to vote for candidate (1-7 scale)", 
               ylab="Density")
```

---

**N = 100**
```{r, fig.height=4}
bootstrap_mean(data=box$box, replicates=10000, N=100, bin=.035,  
               # title="N=100",
               xmin=1, xmax=7,
               xlab="Mean - Respondent wants to vote for candidate (1-7 scale)", 
               ylab="Density")
```



3. Standard error for the difference in means
==========================================================
  
```{r}

#Gerber and Green (2012); Chattopadhyay and Duflo (2004)

# the data
gg_data <- as.data.frame(cbind(c(10,15,20,20,10,15,15), 
                               c(15,15,30,15,20,15,30)))
names(gg_data) <- c("Y_i0", "Y_i1")
```


---

```{r}
# generating empty dataframe to put the results
ate <- as.data.frame(matrix(NA, 10000, 2))
names(ate) <- c("estimated_ate", "estimated_se_ate")

# sampling
for (i in 1:10000){
  
  # generating treatment vector for this replicate
  gg_data$treat <- 0
  gg_data$treat[sample(1:7, 2, replace=F)]  <- 1
  
  treat_mean <- mean(gg_data$Y_i1[gg_data$treat==1])
  treat_var <- var(gg_data$Y_i1[gg_data$treat==1])
  
  control_mean <- mean(gg_data$Y_i0[gg_data$treat==0])
  control_var <- var(gg_data$Y_i0[gg_data$treat==0])
  
  ate[i,1] <- treat_mean - control_mean
  ate[i,2] <- sqrt(treat_var/2 + control_var/5) 
}

```

---

```{r, echo=FALSE}

m <- ggplot(ate, aes(x=estimated_ate))
m + 
  geom_bar(aes(y = 100 * (..count..)/sum(..count..)), binwidth=.5, alpha=.5) + 
  # geom_histogram(aes(y = ..density..)) +
  geom_vline(x=mean(ate$estimated_ate), col="red", size=1.25) +
  theme_bw() +
  xlab("Estimated ATE") +
  ylab("Percent")

```

What should be the title of this figure?


4. Hypothesis testing 
========================================================

```{r}

# generating treatment vector for a given experiment
gg_data$treat <- c(1, 0, 0, 0, 0, 0, 1)

# getting observed outcomes
gg_data$observed <- ifelse(gg_data$treat==1, gg_data$Y_i1, gg_data$Y_i0)

# ate
ATE <- mean(gg_data$observed[gg_data$treat==1]) - mean(gg_data$observed[gg_data$treat==0])
ATE

```

> - What is a p-value?


4a. T-test
========================================================

```{r}
treated <- gg_data$observed[gg_data$treat==1]
treated

var1 <- sum((treated - mean(treated))^2) / (length(treated) - 1)
var1

not_treated <- gg_data$observed[gg_data$treat==0]
not_treated

var0 <- sum((not_treated - mean(not_treated))^2) / (length(not_treated) - 1)
var0

estimated_se <- sqrt(var1/length(treated) + var0/length(not_treated))
estimated_se # Why is this number different from 7.75 (the number in the lecture slides)?
# rounding error!

```

---

```{r}

# converting to standard units
t_stat <- (ATE - 0) / estimated_se
t_stat

# To be able to get the right Student t Distribution, we need to calculate
# the degrees of freedom (Satterthwaite)
df <- (var1/length(treated) + var0/length(not_treated))^2 / 
           ((var1/length(treated))^2 / (length(treated) - 1) + 
           (var0/length(not_treated))^2 / (length(not_treated) - 1))
df

```

---

```{r}

# Overlaying the t_stat to the student t distribution
ggplot(data.frame(x = c(-5, 5)), aes(x)) + 
  stat_function(fun=dt, args=list(df=df, ncp=0), col="blue", size=1) +
  geom_vline(x=mean(t_stat), col="red", size=1.25) 
```

---

```{r}
# One tailed p-value
pt(t_stat, df=df, ncp=0, lower.tail=F)

# Two tailed p-value
pt(-t_stat, df=df, ncp=0, lower.tail=T) + pt(t_stat, df=1.12, ncp=0, lower.tail=F)

```


4b. Randomization inference 
===========================================================================

To get all the possible treatment vectors, we will generate 10000 differente ones...

```{r}
fake_treats <- matrix(NA, 10000, 7)
for (i in 1:10000){
fake_treats[i,] <- sample(gg_data$treat, 7, replace=F)
}

```

... and then only keep the unique ones

```{r}

fake_treats <- unique(fake_treats)

```

---

Now we need to calculate the ATE for each of these possible randomizations. For that, we will 
need a loop

```{r}
rand_ate <- NA # placeholder vector for results

for (i in 1:nrow(fake_treats)){ # for each of the fake treatment vectors
  
  mean_treat <- mean(gg_data$observed[fake_treats[i,]==1])
  
  mean_control <- mean(gg_data$observed[fake_treats[i,]==0])
  
  # calculating ATE for this randomization
  rand_ate[i] <- mean_treat - mean_control
  
}

```

---

Now we can plot the distribution of the randomization ATEs

```{r, echo=FALSE, fig.height=4, fig.cap="Distribution of randomization ATEs"}

m <- ggplot(as.data.frame(rand_ate), aes(x=rand_ate))
m + 
  geom_histogram(aes(y = ..density..), binwidth=.5) +
  geom_vline(x=ATE, col="red", size=1) +
  theme_bw() +
  xlab("Randomization ATEs") +
  ylab("Density")

```

---

And we can get the p-value

```{r}

# One tailed
sum(rand_ate>=ATE)/length(rand_ate)

# Two tailed
sum(abs(rand_ate)>=ATE)/length(rand_ate)


```


